---
title: "Peer-assignment 1"
author: "Kiranmayee Bakshy"
date: "Tuesday, December 09, 2014"
output: html_document
---

### Loading and preprocessing the data


```{r, echo=TRUE}
setwd("C:/Users/kiran/Documents/COURSES-JHU/Reproducible research")
data=read.csv("activity.csv", header=TRUE, sep=",")
```

```{r, echo=TRUE}
#ignore the missing values
complete.data<-complete.cases(data) 
#subset the non-missing values
complete.data<-subset(data,complete.cases(data)==TRUE) 
#split data by date
data.bydate<-split(complete.data,complete.data$date, drop=TRUE)
#calculate the number of steps per day
steps_per_day<-sapply(data.bydate, function(x) sum(x$steps))
#plot histogram
hist(steps_per_day, col="cyan")
#calculate and report the mean and median
mean(steps_per_day)
median(steps_per_day)
```


### What is the average daily activity pattern?

- Time series plot: 5 min interval (x-axis) and average number of steps taken across all days (y-axis)

```{r}
MeanPerInterval <- tapply(complete.data$steps, complete.data$interval, mean)
plot(MeanPerInterval, type = "l", main = ("Steps vs. Interval (daily average)"), ylab = "number of steps")
# The interval with maximum number of steps
seq(along = MeanPerInterval)[MeanPerInterval==max(MeanPerInterval)]
```

### Imputing missing values

- Total number of missing values in the dataset (number of rows with 'NA')

```{r}
sum(is.na(data$steps))
```

- Fill all the missing values with mean of that 5-min interval

```{r}
# Get the steps mean per interval as a vector
tmp_MeanPerInterval <- as.vector(MeanPerInterval)
# Repeat it to be the same for each of the 61 days
tmp_MeanPerInterval <- rep(tmp_MeanPerInterval, 61)
# Set it one where there is no missin data
tmp_MeanPerInterval[!is.na(data$steps)] = 1

# Get the steps data as a vector
tmp_dataTest <- as.vector(data$steps)
# Set it to one where data is missing
tmp_dataTest[is.na(tmp_dataTest)] = 1

# create a copy of the data to be imputed
data_NoMissing <- data
data_NoMissing$steps <- tmp_MeanPerInterval * tmp_dataTest
stepsTotalPerDay_NoMissing <- tapply(data_NoMissing$steps, data_NoMissing$date, 
    sum)
hist(stepsTotalPerDay_NoMissing, breaks = 6, main = "Frequency of number of steps per day", 
    xlab = "steps_per_day", ylab = "Frequency", col = "blue")
```

Mean total number of steps taken per day for the imputed dataset is:

```{r}
mean(stepsTotalPerDay_NoMissing)
```

Median total number of steps taken per day for the imputed dataset is:

```{r}
median(stepsTotalPerDay_NoMissing)
```

- The mean of the new imputed dataset slightly differs from that of the original dataset whereas the median does not. 
- Here, the mean exactly equals the median. 
- From the histogram we can see that the  the number of data in the middle of histogram has increased since number of new data with the mean has been added.

### Differences in activity patterns between weekdays and weekends?

```{r}
# Create a factor variable with two levels (weekday, weekend-day)
tmpLT <- as.POSIXlt(data$date, format = "%Y-%m-%d")
tmpWeekDays <- tmpLT$wday
tmpWeekDays[tmpWeekDays == 0] = 0
tmpWeekDays[tmpWeekDays == 6] = 0
tmpWeekDays[tmpWeekDays != 0] = 1
tmpWeekDaysFactor <- factor(tmpWeekDays, levels = c(0, 1))
# Add the factor variable to the data
data$WD <- tmpWeekDaysFactor
# Calculate the mean
stepsMeanPerWeekday <- tapply(data$steps, list(data$interval, data$WD), mean, 
    na.rm = T)

par(mfrow = c(2, 1))
# Display the 2 plots
with(data, {
    par(mai = c(0, 1, 1, 0))
    plot(stepsMeanPerWeekday[, 1], type = "l", main = ("Steps vs. Interval"), 
        xaxt = "n", ylab = "Weekends")
    title = ("# of Steps v.s. Interval")
    par(mai = c(1, 1, 0, 0))
    plot(stepsMeanPerWeekday[, 2], type = "l", xlab = "Interval", ylab = "Weekdays")

})
```

- Yes, there is clearly a difference in the activity pattern of weekdays and weekends.
